Continuous gravitational waves in the lab: recovering audio signals with a table-top optical microphone

### 2021-05-11 paper/response/AJPRefereeResponse.tex

\subsection*{Comments from R3}

\begin{quote}
The paper describes a table top interferometer for demonstrating the basic operating principles of gravitational wave (GW) detectors. Audio signals are used to demonstrate gravitational wave data analysis techniques for the detection and estimation of signals in noisy data. The paper is well written and will be of interest to educators who wish to provide some hands-on training to students in GW astronomy. I have some comments below that the authors should address.
\end{quote}

\noindent 
-- The data analysis problems discussed in this paper are related to continuous wave searches. However, I did not find a mention of the key feature involved in these searches, namely, Doppler modulation of the signal frequency caused by the rotation of the Earth. This is the principal reason behind the extremely high computational cost of CW searches. A discussion of doppler modulation should be included. (It would be neat if the setup could be extended in the future to show the Doppler effect.)
\begin{itemize}
\item We have including some details on how the detection statistic is used to account for the Earth's motion around the Sun in Section IV A. We have also noted that Doppler modulation of the source also needs to be accounted for when searching for a neutron star in a binary.
\item It would indeed be very nice to modify the setup to show the Doppler effect. We have added some possible ways to explore this topic in our future work discussion in Section VI. 
\item[]
\end{itemize}

\noindent
-- Is the HMM defined properly in Sec. IV A? The observables here are the amplitudes of the spectrogram pixels at a given time index and the hidden markov process is the underlying drift of frequency and amplitude of the signal. I was expecting to see a conditional pdf that connects the state of the underlying (suitably discretized) process to the observables. I don't see it.
\begin{itemize}
\item We have clarified the explanation of the HMM in the text. The detection statistic captures the likelihood of a signal being present in each time-frequency bin conditional on the observed data. In Section IV A we have expanded our description of gravitational wave analysis and referred to both the transmission matrix and emission matrix by name to reduce confusion.% \han{from Andrew: p4, top point - "given the observable data" -> "conditional on the observed data" - and just mention that the conditionality is dropped from the notation [if true] to avoid clutter}
\item[]
\end{itemize}

@@

\noindent
-- Is Eq. 1 correct? The LHS is a marginal probability while the RHS is a joint probability. Shouldn't there be a marginalization on the RHS?
\begin{itemize}
\item This equation was indeed incorrect and we thank you for the careful review. The equation should give the probability for a specific path, and is not a sum over all possibilities. We have corrected equation 1 and clarified what it represents in the text.
\item[]
\end{itemize}

\noindent
-- The normalization used for $F(t_i, f_j)$ should be defined explicitly. 
\begin{itemize}
\item Each value is divided by the maximum value in the grid such that $\max_{i,j} F(t_i, f_j) = 1$. We have added this information to section IV B.
\item[]
\end{itemize}

\noindent
-- Looking at Fig. 3, it appears that the noise is not white and rises in power at lower frequencies. How is the color of the noise taken into account, if at all, in the HMM or the Viterbi algorithm? A discussion of this point should be included although it is fine to assume white noise as a first approximation. 
\begin{itemize}
\item We do indeed assume that the noise is white in our analysis. This assumption is consistent with the use of Viterbi for continuous-wave searches, where white noise is assumed even though LIGO noise is not white%\han{todo get refs for this}. 
We have added some text to clarify this assumption in the first paragraph of Section IV C.
\item[]
\end{itemize}

### 2021-05-11 paper/response/AJPRefereeResponse.tex

\noindent
-- Is the HMM defined properly in Sec. IV A? The observables here are the amplitudes of the spectrogram pixels at a given time index and the hidden markov process is the underlying drift of frequency and amplitude of the signal. I was expecting to see a conditional pdf that connects the state of the underlying (suitably discretized) process to the observables. I don't see it.
\begin{itemize}
\item We have clarified the explanation of the HMM in the text. The detection statistic captures the likelihood of a signal being present in each time-frequency bin conditional on the observed data. In Section IV A we have expanded our description of gravitational wave analysis and referred to both the transmission matrix and emission matrix by name to reduce confusion.% \han{from Andrew: p4, top point - "given the observable data" -> "conditional on the observed data" - and just mention that the conditionality is dropped from the notation [if true] to avoid clutter}
\item[]
\end{itemize}

\noindent
-- Is Eq. 1 correct? The LHS is a marginal probability while the RHS is a joint probability. Shouldn't there be a marginalization on the RHS?
\begin{itemize}
\item This equation was indeed incorrect and we thank you for the careful review. The equation should give the probability for a specific path, and is not a sum over all possibilities. We have corrected equation 1 and clarified what it represents in the text.
\item[]
\end{itemize}

\noindent
-- The normalization used for $F(t_i, f_j)$ should be defined explicitly. 
\begin{itemize}
\item Each value is divided by the maximum value in the grid such that $\max_{i,j} F(t_i, f_j) = 1$. We have added this information to section IV B.
\item[]
\end{itemize}

\noindent
-- Looking at Fig. 3, it appears that the noise is not white and rises in power at lower frequencies. How is the color of the noise taken into account, if at all, in the HMM or the Viterbi algorithm? A discussion of this point should be included although it is fine to assume white noise as a first approximation. 
\begin{itemize}
\item We do indeed assume that the noise is white in our analysis. This assumption is consistent with the use of Viterbi for continuous-wave searches, where white noise is assumed even though LIGO noise is not white (for example, see Suvorova et al. Phys. Rev. D, Vol 96, Issue 10, 102006 (2017). We have added some text to clarify this assumption in the first paragraph of Section IV C.
\item[]
\end{itemize}

### 2021-05-11 paper/response/AJPRefereeResponse.tex

\subsection*{Comments from R3}

\begin{quote}
The paper describes a table top interferometer for demonstrating the basic operating principles of gravitational wave (GW) detectors. Audio signals are used to demonstrate gravitational wave data analysis techniques for the detection and estimation of signals in noisy data. The paper is well written and will be of interest to educators who wish to provide some hands-on training to students in GW astronomy. I have some comments below that the authors should address.
\end{quote}

\noindent 
-- The data analysis problems discussed in this paper are related to continuous wave searches. However, I did not find a mention of the key feature involved in these searches, namely, Doppler modulation of the signal frequency caused by the rotation of the Earth. This is the principal reason behind the extremely high computational cost of CW searches. A discussion of doppler modulation should be included. (It would be neat if the setup could be extended in the future to show the Doppler effect.)
\begin{itemize}
\item We have including some details on how the detection statistic is used to account for the Earth's motion around the Sun in Section IV A. We have also noted that Doppler modulation of the source also needs to be accounted for when searching for a neutron star in a binary.
\item It would indeed be very nice to modify the setup to show the Doppler effect. We have added some possible ways to explore this topic in our future work discussion in Section VI. 
\item[]
\end{itemize}

\noindent
-- Is the HMM defined properly in Sec. IV A? The observables here are the amplitudes of the spectrogram pixels at a given time index and the hidden markov process is the underlying drift of frequency and amplitude of the signal. I was expecting to see a conditional pdf that connects the state of the underlying (suitably discretized) process to the observables. I don't see it.
\begin{itemize}
\item We have clarified the explanation of the HMM in the text. The detection statistic $F(t_i, f_j)$ captures the likelihood of a signal being present in each time-frequency bin conditional on the observed data, where we have dropped the conditionally from the notation to avoid clutter. In Section IV A we have expanded our description of gravitational wave analysis and referred to both the transmission matrix and emission matrix by name to reduce confusion.
\item[]
\end{itemize}

### 2021-05-20 paper/paper_main.tex
getting manuscript ready for resubmission (removing identifying information for double anoymous review)

%\author{\textbf{[Authors removed for double-anonymous review.]}}

@@

\author{Changrong Liu}
\affiliation{%
Department of Electrical and Electronic Engineering, University of Melbourne, Parkville, Victoria 3010, Australia
}

@@

\begin{abstract}
Gravitational-wave observatories around the world are searching for continuous waves: persistent signals from sources such as spinning neutron stars. 
These searches use sophisticated statistical techniques to look for weak signals in noisy data. 
In this paper, we demonstrate these techniques using a table-top model gravitational-wave detector: a Michelson interferometer where sound is used as an analog for gravitational waves. 
Using signal processing techniques from continuous-wave searches, we demonstrate the recovery of tones with constant and wandering frequencies. 
We also explore the use of the interferometer as a teaching tool for educators in physics and electrical engineering by using it as an ``optical microphone'' to capture music and speech. 
% \jam{We mention undergraduate labs twice, roll this sentence into the end? Also, electrical sounds weird, why the change from electrical engineering?}
%The interferometer is used as an ``optical microphone'' to capture music and speech. 
% \jam{I changed all the \emph to ``'', feel free to change them back, but I think it is more explicit this way}
A range of filtering techniques used to recover signals from noisy data are detailed in the Supplementary Material. 
Here, we present highlights of our results using a combined notch plus Wiener filter and the statistical log minimum mean-square error (logMMSE) estimator. 
%We present a range of example filtering techniques to recover the original signal from noisy data, including a bandpass filter, notch filters, a Wiener filter, and the statistical log minimum mean-square-error (logMMSE) estimator.
% log minimum mean-square-error (logMMSE) estimator
Recordings of simple chords and drums are easily recovered using these techniques, but complex music and speech are more challenging. 
This demonstration can be used by educators in undergraduate laboratories and can be adapted for use as an engagement tool for communicating gravitational-wave and signal-processing topics to non-specialist audiences. 
% This demonstration has scope for use by educators in undergraduate laboratories and can be adapted for use as an engagement tool for communicating gravitational-wave and signal-processing topics to non-specialist audiences. 
% rearranged in response to Anna's pnp review
%Finally we use the interferometer as an `optical microphone' to capture music and speech and apply filtering techniques to recover the original signals from the noisy data. Such techniques include simple bandpass filtering, notch filters, a Weiner filter, and the statistical logMMSE method.  
%Simple chords and drums are easily recovered, but complex music and speech are more challenging. 
%This demonstration has scope for use in physics and electrical engineering undergraduate laboratories. 
%It can also be adapted for use as an engagement tool for communicating gravitational wave and signal-processing topics to non-specialist audiences. 
\end{abstract}

@@

\begin{acknowledgments}
%\textbf{[Removed for double-anonymous review.]} 
The authors are grateful to Jude Prezens, Alex Tolotchkoc, and Blake Molyneux for their technical advice and generous assistance; Patrick Meyers, Margaret Millhouse, Sofia Suvorova for useful discussions; Patrick Clearwater, Patrick Meyers, Suk Yee Yong, Lucy Strang, Julian Carlin, Sanjaykumar Patil, and Alex Cameron for early work on the interferometer design requirements and construction; and the LIGO Education Public Outreach working group, in particular, Anna Green, Lynn Cominsky, and Sam Cooper, for their helpful feedback and suggestions.  
This research is supported by the Australian Research Council Centre of Excellence for Gravitational Wave Discovery (OzGrav) (project number CE170100004). 
Financial support towards hardware was provided by the Institute of Physics International Member Grant and the OzGrav Outreach Support scheme. 
Travel support was provided by the Australian National University PhB Science program.
This work has been assigned LIGO document number P2000386.

>>>>

\author{\textbf{[Authors removed for double-anonymous review.]}}

\begin{comment}

@@

\author{Changrong Liu}
\affiliation{%
Department of Electrical and Electronic Engineering, University of Melbourne, Parkville, Victoria 3010, Australia
}
\affiliation{%
OzGrav-Melbourne, Australian Research Council Centre of Excellence for Gravitational Wave Discovery, Parkville, Victoria 3010, Australia
}

@@

\begin{abstract}
Gravitational-wave observatories around the world are searching for continuous waves: persistent signals from sources such as spinning neutron stars. 
These searches use sophisticated statistical techniques to look for weak signals in noisy data. 
In this paper, we demonstrate these techniques using a table-top model gravitational-wave detector: a Michelson interferometer where sound is used as an analog for gravitational waves. 
Using signal processing techniques from continuous-wave searches, we demonstrate the recovery of tones with constant and wandering frequencies. 
We also explore the use of the interferometer as a teaching tool for educators in physics and electrical engineering by using it as an ``optical microphone'' to capture music and speech. 
A range of filtering techniques used to recover signals from noisy data are detailed in the Supplementary Material. 
Here, we present highlights of our results using a combined notch plus Wiener filter and the statistical log minimum mean-square error (logMMSE) estimator. 
Recordings of simple chords and drums are easily recovered using these techniques, but complex music and speech are more challenging. 
This demonstration can be used by educators in undergraduate laboratories and can be adapted for use as an engagement tool for communicating gravitational-wave and signal-processing topics to non-specialist audiences.  
\end{abstract}

@@

\begin{acknowledgments}
\textbf{[Removed for double-anonymous review.]}

\begin{comment}
The authors are grateful to Jude Prezens, Alex Tolotchkoc, and Blake Molyneux for their technical advice and generous assistance; Patrick Meyers, Margaret Millhouse, Sofia Suvorova for useful discussions; Patrick Clearwater, Patrick Meyers, Suk Yee Yong, Lucy Strang, Julian Carlin, Sanjaykumar Patil, and Alex Cameron for early work on the interferometer design requirements and construction; and the LIGO Education Public Outreach working group, in particular, Anna Green, Lynn Cominsky, and Sam Cooper, for their helpful feedback and suggestions.  
This research is supported by the Australian Research Council Centre of Excellence for Gravitational Wave Discovery (OzGrav) (project number CE170100004). 
Financial support towards hardware was provided by the Institute of Physics International Member Grant and the OzGrav Outreach Support scheme. 
Travel support was provided by the Australian National University PhB Science program.
This work has been assigned LIGO document number P2000386.
\end{comment}
